---
title: "Practical Machine Learning Course Project"
author: "Chirag Soni"

---

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```

---
## Project Details

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har 

### Goal

The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


### Data

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


### Getting and loading the data

```{r}
set.seed(12345)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))

**Lets check the dimensions of these tables:**

dim(training)
19622   160


dim(testing)
20 160


summary(training$classe)
   A    B    C    D    E 
5580 3797 3422 3216 3607
```
**Partioning the training set into two**

train_input <- createDataPartition(training$classe, p=0.6, list=FALSE)
training_data <- training[train_input, ]
testing_data <- training[-train_input, ]
dim(training_data)
 [1] 11776   160


dim(testing_data)
 [1] 7846  160
```


**Cleaning the data**

**In this step we clean the data, remove near 0 variance variables which are anyways of no use in modeling**
```{r}
nzv <- nearZeroVar(training_data, saveMetrics=TRUE)
training_data <- training_data[,nzv$nzv==FALSE]



dim(training_data)
 [1] 11776   133



nzv<- nearZeroVar(testing_data,saveMetrics=TRUE)
testing_data <- testing_data[,nzv$nzv==FALSE]


dim(testing_data)
 [1] 7846  130


training_data <- training_data[c(-1)]


```
**We need to remove variables that have >60% null values**

```{r}
training_new <- training_data
for(i in 1:length(training_data)) {
    if( sum( is.na( training_data[, i] ) ) /nrow(training_data) >= .7) {
        for(j in 1:length(training_new)) {
            if( length( grep(names(training_data[i]), names(training_new)[j]) ) == 1)  {
                training_new <- training_new[ , -j]
            }   
        } 
    }
}



dim(training_new)
## [1] 11776    58



training_data <- training_new
rm(training_new)

```
** Here we transform the training and testing data sets**
```{r}
clean1 <- colnames(training_data)
clean2 <- colnames(training_data[, -58])  # remove the classe column
testing_data <- testing_data[clean1]         # allow only variables in testing_data that are also in training_data
testing <- testing[clean2]             # allow only variables in testing that are also in training_data

dim(testing_data)
## [1] 7846   58
dim(testing)
## [1] 20 57



**Coerce the data into the same type**

for (i in 1:length(testing) ) {
    for(j in 1:length(training_data)) {
        if( length( grep(names(training_data[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(training_data[i])
        }      
    }      
}

**To get the same class between testing and training_data**
testing <- rbind(training_data[2, -58] , testing)
testing <- testing[-1,]



**Hee is the model equation, using rpart to create decision trees:**
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=training_data, method="class")
```

**Use the model created above to predict test data**

```{r}
predictionsA1 <- predict(modFitA1, testing_data, type = "class")
cmtree <- confusionMatrix(predictionsA1, testing_data$classe)
cmtree
## Confusion Matrix and Statistics
##
##          Reference
## Prediction    A    B    C    D    E
##         A 2150   60    7    1    0
##         B   61 1260   69   64    0
##         C   21  188 1269  143    4
##         D    0   10   14  857   78
##         E    0    0    9  221 1360
##
## Overall Statistics
##                                          
##               Accuracy : 0.8789          
##                 95% CI : (0.8715, 0.8861)
##    No Information Rate : 0.2845          
##    P-Value [Acc > NIR] : < 2.2e-16       
##                                          
##                  Kappa : 0.8468          
## Mcnemar's Test P-Value : NA              

## Statistics by Class:

##                     Class: A Class: B Class: C Class: D Class: E
## Sensitivity            0.9633   0.8300   0.9276   0.6664   0.9431
## Specificity            0.9879   0.9693   0.9450   0.9845   0.9641
## Pos Pred Value         0.9693   0.8666   0.7809   0.8936   0.8553
## Neg Pred Value         0.9854   0.9596   0.9841   0.9377   0.9869
## Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
## Detection Rate         0.2740   0.1606   0.1617   0.1092   0.1733
## Detection Prevalence   0.2827   0.1853   0.2071   0.1222   0.2027
## Balanced Accuracy      0.9756   0.8997   0.9363   0.8254   0.9536
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
	```


**Try the prediction with Random Forests**
```{r}
set.seed(1234)
modFitB1 <- randomForest(classe ~ ., data=training_data)
predictionB1 <- predict(modFitB1, testing_data, type = "class")
cmrf <- confusionMatrix(predictionB1, testing_data$classe)
cmrf
## Confusion Matrix and Statistics
##
##          Reference
## Prediction    A    B    C    D    E
##         A 2231    2    0    0    0
##         B    1 1516    2    0    0
##         C    0    0 1366    3    0
##         D    0    0    0 1281    2
##         E    0    0    0    2 1440
##
## Overall Statistics
##                                          
##               Accuracy : 0.9985          
##                 95% CI : (0.9973, 0.9992)
##    No Information Rate : 0.2845          
##    P-Value [Acc > NIR] : < 2.2e-16       
##                                          
##                  Kappa : 0.9981          
## Mcnemar's Test P-Value : NA              
##
## Statistics by Class:
##
##                     Class: A Class: B Class: C Class: D Class: E
## Sensitivity            0.9996   0.9987   0.9985   0.9961   0.9986
## Specificity            0.9996   0.9995   0.9995   0.9997   0.9997
## Pos Pred Value         0.9991   0.9980   0.9978   0.9984   0.9986
## Neg Pred Value         0.9998   0.9997   0.9997   0.9992   0.9997
## Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
## Detection Rate         0.2843   0.1932   0.1741   0.1633   0.1835
## Detection Prevalence   0.2846   0.1936   0.1745   0.1635   0.1838
## Balanced Accuracy      0.9996   0.9991   0.9990   0.9979   0.9992


plot(modFitB1)


plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

**We will try to fit the GBM model as per below:**
```{r}
set.seed(1234)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmFit1 <- train(classe ~ ., data=training_data, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)


gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=testing_data)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, testing_data$classe)
gbmAccuracyTest
## Confusion Matrix and Statistics
## 
##           Reference
## Prediction    A    B    C    D    E
##          A 2232    4    0    0    0
##          B    0 1513    1    0    0
##          C    0    1 1362    4    0
##          D    0    0    5 1273    1
##          E    0    0    0    9 1441
## 
## Overall Statistics
 ##                                          
##                Accuracy : 0.9968          
##                  95% CI : (0.9953, 0.9979)
##     No Information Rate : 0.2845          
##     P-Value [Acc > NIR] : < 2.2e-16       
 ##                                          
##                   Kappa : 0.996           
##  Mcnemar's Test P-Value : NA              

## Statistics by Class:

##                      Class: A Class: B Class: C Class: D Class: E
## Sensitivity            1.0000   0.9967   0.9956   0.9899   0.9993
## Specificity            0.9993   0.9998   0.9992   0.9991   0.9986
## Pos Pred Value         0.9982   0.9993   0.9963   0.9953   0.9938
## Neg Pred Value         1.0000   0.9992   0.9991   0.9980   0.9998
## Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
## Detection Rate         0.2845   0.1928   0.1736   0.1622   0.1837
## Detection Prevalence   0.2850   0.1930   0.1742   0.1630   0.1848
## Balanced Accuracy      0.9996   0.9983   0.9974   0.9945   0.9990


plot(gbmFit1, ylim=c(0.9, 1))
```

**We get better accuracy on using random forest as compared to GBM**
```{r}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
## 1  2 31  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 
## B  A  B  A  A  E  D  B  A  A  B  C  B  A  E  E  A  B  B  B 
## Levels: A B C D E

```
